---
title: "`evoland-plus` tutorial"
author: Jan Hartman
date: last-modified
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
number-sections: true
---

```{r}
#| label: cleanup
#| include: false
# TODO render tables into nice HTML interactive ones??

# make sure we're starting clean
unlink("firstmodel.evolanddb", recursive = TRUE)

local({
  chunk_messages <- character()

  knitr::knit_hooks$set(message = function(x, options) {
    chunk_messages <<- c(chunk_messages, x)
    return(character(0))
  })

  default_chunk_hook <- knitr::knit_hooks$get("chunk")
  knitr::knit_hooks$set(chunk = function(x, options) {
    out <- default_chunk_hook(x, options)
    if (length(chunk_messages) > 0) {
      msg_text <- paste(chunk_messages, collapse = "")
      callout <- paste0(
        "\n\n<details><summary>Messages</summary><pre>\n",
        msg_text,
        "</pre></details>\n"
      )
      chunk_messages <<- character()
      out <- paste0(out, callout)
    }
    return(out)
  })
})
```

*Note: This tutorial expects that you already know about geodata and some central principles of pattern-based land use change modelling.*

This file introduces the `evoland-plus` R package, or `evoland` for short.
`evoland-plus` builds on the idea that land use change can be predicted from observed change patterns.
The likelihood that a location and its neighbors will change state is called *transition potential* and is estimated per location.
A concrete realization of all possible transitions (e.g. forest->grass, grass->forest, etc.) is realized in an *allocation* across all locations and classes.
Because the transition potential is spatially and temporally autocorrelated, an extrapolation must be autoregressive.

# Workflow {.unnumbered}

```{mermaid}
---
config:
    flowchart:
        useMaxWidth: true
---
flowchart TD
    subgraph setup[1. Setup]
        direction LR
        domain_categories{{ fa:fa-table-list LULC Categorization}}
        domain_space{{ fa:fa-globe Spatial Domain }}
        domain_time{{ fa:fa-clock Temporal Domain}}
        domain_categories --- domain_space
        domain_space --- domain_time
    end
    setup --> preparation

    subgraph preparation["2. Ingesting Data"]
        direction LR
        predictors@{shape : docs, label: "Spatiotemporal Predictors"}
        lulc_data@{shape : doc, label: " LULC data \n @ {t, t-1, t-2, ...}"}
        ts@{shape : brace, label: " Timesteps: \n t = now \n t-1 = one step in the past \n t+1 = one step in the future"}
        style ts stroke:#ccc, stroke-width:2px
        predictors --- lulc_data
        lulc_data --- ts
    end
    preparation --> calibration


    subgraph calibration["3. Calibration"]
        direction LR
        varsel[Select Features]
        varsel --> markovmods
        markovmods["Train Transition<br> Potential Models"]
        markovmods --> parametrize_alloc
        parametrize_alloc["Estimate Allocation Parameters"]
    end

    calibration --> estimation

    subgraph estimation["4. Prediction + Allocation"]
        direction LR
        evalmodel["`Evaluate Transition Poten Models @ t`"]
        predictions["Transition Potential Maps @ t+1"]
        evalmodel --> predictions
        changes["Allocate projected land use demand <br> (patch / expand)"]
        predictions --> changes
    end

    estimation --> projection
    projection@{shape : terminal, label: "LULC projection\n @ t+1"}
    projection -->|Next iteration with t+1 as t| estimation

    linkStyle 0,1,3,4 stroke-opacity:0

    classDef phase fill:#f9f9f990,stroke:#333,stroke-dasharray:2 4,color:#000000
    classDef data fill:#ddf1d5,stroke:#82b366,color:#000000
    classDef user_input fill:#fff2cc,stroke:#d6b655,color:#000000

    class setup,calibration,preparation,estimation,allocation phase
    class predictors,lulc_data,projection data
    class domain_categories,domain_space,domain_time user_input
```

# Setup

We'll use the following packages:

```{r}
#| label: setup
#| output: false
library(evoland)
library(data.table)
library(terra)
```

## Creating an `evoland` database

First, we load the package and create an `evoland_db` database object: unlike most R objects (e.g. `data.frame`), this is a mutable object[^r6], meaning we can alter its state like we would with a Python object.
This nicely mirrors our persistent database on disk.

[^r6]: Specifically, we're creating an R6 object instead of an S3 or S4 one. This is called "encapsulated object oriented programming" (cf. [Advanced R](https://adv-r.hadley.nz/r6.html)) and you might know it from Python (e.g. `pandas_df.drop_duplicates()`).

If there is no directory at `path`, we'll create a new DB: a set of parquet files following clearly defined parquet files.
If there _is_ a directory, we resume from where we left off.

```{r}
db <- evoland_db$new(path = "firstmodel.evolanddb")
```

Go ahead and print the `d`` object.
There are already a `runs_t` and a `reporting_t` table, which are bare-bones for now but will be used to track our [modelling runs](../reference/runs_t.html) and metadata for producing graphs and tables.

```{r}
#| label: print-db
db
```

## Defining our model framework

Before we do anything else, let's declare the framework of our model:

- **Land Use / Cover Categories**: A set of *land use and/or land cover classes* between which changes can occur
- **Spatial Domain**: A set of *coordinates*, built from a geographic specification
- **Temporal Domain**: A set of *periods*, describing a regular time series

### LULC Categories

We define our fundamental LULC classes. The `src_classes` field maps the underlying data source's classes to our conceptual categories.

```{r}
#| label: recode-arealstatistik
db$lulc_meta_t <- create_lulc_meta_t(
  list(
    forest = list(
      pretty_name = "Forest",
      description = "Areas with lots of trees",
      src_classes = 1:3
    ),
    arable = list(
      pretty_name = "Arable Land",
      src_classes = c(4, 8)
    ),
    urban = list(
      pretty_name = "Urban Areas",
      description = "Where nature goes to die",
      src_classes = 5:7
    ),
    static = list(
      pretty_name = "Immutable",
      description = "Areas where we cannot conceptualize change",
      src_classes = 9:10
    )
  )
)
```

### Spatial domain

The spatial coordinate points provide the fundamental geographic domain on which our model will operate.
We use a dense square raster, but since we register individual coordinate points, we could subset these to any oddly shaped region of interest.

```{r}
#| label: set-coords
# template SpatRaster: 30x30 grid in Swiss LV95, later used for synthetic data generation
template_rast <- terra::rast(
  crs = "EPSG:2056",
  extent = terra::ext(c(
    xmin = 2697000,
    xmax = 2700000,
    ymin = 1252000,
    ymax = 1255000
  )),
  resolution = 100
)

db$coords_t <- create_coords_t_square(
  epsg = terra::crs(template_rast, describe = TRUE)$code |> as.integer(),
  extent = terra::ext(template_rast),
  resolution = terra::res(template_rast)[1]
)
```

You can retrieve coords_t from disk and filter using [`data.table` semantics](https://raw.githubusercontent.com/rstudio/cheatsheets/master/datatable.pdf):

```{r}
db$coords_t[lon == 2699650]
```

We can also retrieve a minimal representation (id, lat, lon) using an active binding, i.e. a method tied to the database that dynamically computes a property when called:

```{r}
db$coords_minimal[1:2]
```

### Temporal domain

We also define our temporal domain.
Note that an additional "0th" period is added at the end of the observed range; this is used for labelling predictor or intervention data as static.

```{r}
#| label: set-periods
db$periods_t <- create_periods_t(
  period_length_str = "P10Y", # 10 year period
  start_observed = "1995-01-01",
  end_observed = "2020-01-01",
  end_extrapolated = "2030-01-01"
)
```

# Ingesting Data

## LULC Data

For this tutorial, we do not read real observed LULC data:
instead, we generate a synthetic LULC raster with 3 layers (one per period) by adding autoregressive noise to a noisy raster.

```{r}
#| label: synthesize-and-agg-lulc-data
#| fig-asp: 0.3
# autoregressive noise with skellam distribution
n_cells <- dim(template_rast)[1] * dim(template_rast)[2]
noise1 <- runif(n_cells, min = 0, max = 10)
noise2 <- noise1 + stats::rpois(n_cells, 1) - stats::rpois(n_cells, 1)
noise3 <- noise2 + stats::rpois(n_cells, 1) - stats::rpois(n_cells, 1)

synthetic_lulc <-
  rast(template_rast, nlyrs = 3, vals = c(noise1, noise2, noise3)) |>
  focal(w = 3, fun = mean, na.rm = TRUE) |>
  clamp(lower = 0, upper = 10) |>
  classify(rcl = data.frame(from = 0:9, to = 1:10, becomes = sample(1:10, 10)))

plot(synthetic_lulc, nc = 3)
```

We now extract the generated values at our coordinates using `extract_using_coords_t`, giving us a tabular representation of `id_coord, id_period, src_class` tuples.
In a second step, we join in a long representation of the LULC metadata, associating `id_lulc` with `src_class`.

```{r}
synthetic_at_coords <- extract_using_coords_t(synthetic_lulc, db$coords_t)

synthetic_joint_meta <-
  synthetic_at_coords[, .(
    id_coord,
    id_period = substr(layer, 4, 4) |> as.integer(),
    src_class = value
  )][
    db$lulc_meta_long_v, # map from id_lulc to src_class
    on = .(src_class),
    nomatch = NULL
  ]
```

We now have the LULC data in an almost canonical format.
We need to add information on which `id_run` this data belongs to: run 0 is the base run, providing observed data (see [`db$runs_t`](../reference/runs_t.html) for details).
The call to [`as_lulc_data_t`](../reference/lulc_data_t.html) ensures that the data we want to insert actually conforms to the form we are expecting.

```{r}
db$lulc_data_t <- as_lulc_data_t(synthetic_joint_meta[, .(
  id_run = 0L, # Base run ID
  id_period,
  id_lulc,
  id_coord
)])
```

Now that we have LULC data in our DB, we can derive data from it, e.g. we grab a view where a transition occurred, i.e. the anterior and posterior LULC ID are not the same.

```{r}
db$trans_v[id_lulc_anterior != id_lulc_posterior]
```

## Add Predictors and Neighbors

You have seen above how [`extract_using_coords_t`](../reference/util_terra.html) can be used to transform a `SpatRaster` into a tabular form.
It can also be used with `SpatVector` objects, and the resulting tables need not just be LULC data: we could also use it to extract predictor information.
For demo purposes, we'll use test data that comes with `evoland`.
Note that the metadata holds a `fill_value` used to define what value should be used at a coordinate point with no explicit data set - e.g. if a square kilometre does not have a population count set, we can infer that it should be zero.

```{r}
#| label: add-predictors
db$pred_meta_t <- evoland:::test_pred_meta_t
db$pred_data_t <- evoland:::test_pred_data_t
```

Statistical models like GLMs or random forests lack inherent spatial concepts, so we explicitly calculate neighborhood relations for each coordinate to create spatial predictors (e.g., "number of neighboring forest cells within 100m").
This is done by creating a neighbor lookup table ([`db$set_neighbors`](../reference/neighbors_t.html) and then counting the number of neighbors within each distance break class and land use category (`db$generate_neighbor_predictors`).

```{r}
#| label: neighbors
#| out: false
db$set_neighbors(
  max_distance = 1000,
  distance_breaks = c(0, 100, 500, 1000),
  quiet = TRUE
)

db$generate_neighbor_predictors()
```

Have a look at the predictor metadata we have defined, it now contains new rows for the neighbor predictors:

```{r}
db$pred_meta_t
```

# Calibration

## Eligible Transitions and Predictor Pruning

We filter transitions eligible for modeling based on a minimum number of observed occurrences.

```{r}
#| label: eligible-transitions-and-predictors
db$trans_meta_t <- create_trans_meta_t(db$trans_v, min_cardinality_abs = 50)
```

Because each transition may be modelled using different predictors (aka *features*), we start out by setting the `trans_preds_t` table to the full cross product of viable transitions and predictors.
We then carry out a feature selection step in `get_pruned_trans_preds_t`, which here is used with a two-stage covariance filter.
First, candidate features are ranked and then selected up to a given correlation threshold, see [`covariance_filter`](../reference/covariance_filter.html).
The assignment to `db$trans_preds_t` will overwrite the existing relations; you will be prompted if you want to do this if you're running R interactively.

```{r}
#| label: prune-trans-preds
#| output: false
db$set_full_trans_preds()
# Note: we're suppressing warnings about non-convergence due to our random synthetic data.
db$trans_preds_t <- db$get_pruned_trans_preds_t(
  filter_fun = covariance_filter,
  corcut = 0.1
)
```

## Transition Models

Now we fit partial models using training/validation splits, allowing for a goodness-of-fit (gof) estimation.
Here, we only fit a single partial model per transition, where we normally would create a list of candidates.
Next, fit_full_models() reads the partial models stored in db$trans_models_t, chooses the best model for each transition based on goodness of fit, and refits each chosen model on all available predictor data. Assigning the result back to db$trans_models_t stores these full models for the extrapolation step.

```{r}
#| label: trans-models
db$trans_models_t <- db$fit_partial_models(
  learner = mlr3::lrn("classif.featureless", predict_type = "prob"),
  measures = c("classif.auc", "classif.acc"),
  sample_frac = 0.7,
  seed = 42
)

db$trans_models_t <- db$fit_full_models(
  gof_criterion = "auc",
  gof_maximize = TRUE
)
```

## Transition Rates and Allocation Parameters

As a constrained pattern-based model, we need to provide transition rates to the DinamicaEGO allocator.
In a simple approach, we can extrapolate the rate of each transition from the observed data:

```{r}
#| label: transition-rates-alloc-params
db$trans_rates_t <-
  db$get_obs_trans_rates() |>
  extrapolate_trans_rates(
    periods = db$periods_t,
    coord_count = n_cells
  )
```

We estimate allocation parameters for DinamicaEGO, which determine the shape and size of new patches, respectively which fraction of converted land use is in new versus expanded patches.
This estimation procedure is not unbiased and hence a single estimate may not be enough:
normally, we would perturb the estimate and use multiple `id_run`s to identify the best parametrization.
For simplicity, we now just take the estimates for granted and assign `id_run=0`, i.e. the base run ID.

```{r}
alloc_for_eval <- db$create_alloc_params_t(n_perturbations = 0)
alloc_for_eval[, id_run := 0L] # overwrite id_run=1
db$alloc_params_t <- alloc_for_eval
```

# Prediction + Allocation

With all components in place, we run the allocation step.
If Dinamica is not installed, you'll get a warning that the anterior LULC map is returned as the posterior.
For installation instructions, see the [Installing Dinamica EGO](install-dinamica.html) vignette.

```{r}
#| label: allocation
#| results: hold
db$alloc_dinamica(
  id_period = db$periods_t[is_extrapolated == TRUE, id_period],
  gof_criterion = "auc",
  gof_maximize = TRUE
)
```

## Visualization

Finally, we can extract the simulated LULC maps into `SpatRaster` objects to visualize them.

```{r}
#| label: visualization
labels <- db$periods_t[
  id_period != 0,
  paste0(year(start_date), " to ", year(end_date))
]
plot_maps <- db$lulc_data_as_rast() |> setNames(labels)

plot(plot_maps, type = "classes", levels = db$lulc_meta_t$pretty_name)
```

Of the four maps, only the first three show changes.
Due to Dinamica not being available on our github runner, the extrapolated step should look exactly like the step before.
