---
title: "`evoland-plus` tutorial"
author: Jan Hartman
date: last-modified
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
number-sections: true
---


```{r}
#| label: cleanup
#| include: false
# TODO cleanup output (chunks with newlines in single html parag)
# TODO cleanup output suppress warnings
# TODO add labels to final plot for LULC IDs
# TODO render tables into nice HTML interactive ones??

# make sure we're starting clean
unlink("firstmodel.evolanddb", recursive = TRUE)
```

*Note: This tutorial expects that you already know about geodata and some central principles of pattern-based land use change modelling.*

This file introduces the `evoland-plus` R package, or `evoland` for short.
`evoland-plus` builds on the idea that the land use change can be predicted from observed change patterns.
The *transition potential* from each land use class to every other is estimated per location.
Then, a shift in land use demand is *allocated* across all locations and classes.
The allocation is a *realization* of a stochastic process.

In this tutorial, we'll follow this workflow:

```{mermaid}
---
config:
  theme: base
  fontSize: 20px
  themeVariables:
    cScale0: "lightgrey"
    cScaleInv0: "black"
    cScale1: "#aec7e8"
    cScaleInv1: "#1f77b4"
    cScale2: "#ff9896"
    cScaleInv2: "#d62728"
---

timeline
    section Preparation
    1. Setup
        : Create DB
        : Define LULC categories
        : Define spatial domain
        : Define temporal domain

    2. Data Ingestion
        : Ingest LULC data for past and present timesteps
        : Ingest spatiotemporal predictor data

    section Calibration
    3. Calibration Phase
        : Select features for modeling
        : Train transition potential models
        : Estimate allocation parameters

    section Estimation + Allocation
    4. Prediction & Allocation
        : Evaluate transition potential models at current timestep (t)
        : Generate transition potential maps for next timestep (t+1)
        : Allocate projected land use demand (patch / expand)

    5. Autoregressive extrapolation
        : Write LULC projection for t+1
        : Recalculate neighborhood predictors
        : Back to step
```

# Setup

We'll use the following packages:

```{r}
#| label: setup
#| echo: false
library(evoland)
library(data.table)
library(terra)
```

## Creating an `evoland` database

First, we load the package and create a `evoland_db` database object: unlike most R objects like `data.frame`s, this is a **mutable** object[^r6], meaning we can alter its state without creating a copy.
This nicely mirrors a durable database on disk.

[^r6]: Specifically, we're creating an R6 object instead of an S3 or S4 one. This is called "encapsulated object oriented programming" (cf. [Advanced R](https://adv-r.hadley.nz/r6.html)) and you might know it from Python (e.g. `pandas_df.drop_duplicates()`).

Upon creation, the `path` argument determines where we'll store our data, or if there is already a database at that location, we can continue working with it.

```{r}
db <- evoland_db$new(path = "firstmodel.evolanddb")
```

Go ahead and print the object: there are already a `runs_t` table and a `reporting_t` table, which are bare-bones for now but will be used to track our [modelling runs](reference/runs_t.html) and metadata for producing graphs and tables.

```{r}
#| label: print-db
db
```

## Defining our model framework

Before we do anything else, let's declare the framework of our model:

- **Land Use / Cover Categories**: A set of *land use and/or land cover classes* between which changes can occur
- **Spatial Domain**: A set of *coordinates*, built from a geographic specification
- **Temporal Domain**: A set of *periods*, describing a regular time series

### LULC Categories

We define our fundamental LULC classes. The `src_classes` field maps the underlying data source's classes to our conceptual categories.

```{r}
#| label: recode-arealstatistik
db$lulc_meta_t <- create_lulc_meta_t(
  list(
    forest = list(
      pretty_name = "Forest",
      description = "Areas with lots of trees",
      src_classes = 1:3
    ),
    arable = list(
      pretty_name = "Arable Land",
      src_classes = c(4, 8)
    ),
    urban = list(
      pretty_name = "Urban Areas",
      description = "Where nature goes to die",
      src_classes = 5:7
    ),
    static = list(
      pretty_name = "Immutable",
      description = "Areas where we cannot conceptualize change",
      src_classes = 9:10
    )
  )
)
```

### Spatial domain

The spatial coordinate points provide the fundamental geographic domain on which our model will operate.
We use a dense square raster, but since we register individual coordinate points, we could subset these to any oddly shaped region of interest.

```{r}
#| label: set-coords
# template SpatRaster: 30x30 grid in Swiss LV95, later used for synthetic data generation
template_rast <- terra::rast(
  crs = "EPSG:2056",
  extent = terra::ext(c(
    xmin = 2697000,
    xmax = 2700000,
    ymin = 1252000,
    ymax = 1255000
  )),
  resolution = 100
)

db$coords_t <- create_coords_t_square(
  epsg = terra::crs(template_rast, describe = TRUE)$code |> as.integer(),
  extent = terra::ext(template_rast),
  resolution = terra::res(template_rast)[1]
)
```

You can retrieve coords_t from disk and filter using data.table semantics:

```{r}
db$coords_t[lon == 2699650]
```

We can also retrieve a minimal representation (id, lat, lon) using an active binding:

```{r}
db$coords_minimal
```

### Temporal domain

We also define our temporal domain.
Note that an additional "0th" period is added at the end of the observed range; this is used for labelling predictor or intervention data as static.

```{r}
#| label: set-periods
db$periods_t <- create_periods_t(
  period_length_str = "P10Y", # 10 year period
  start_observed = "1995-01-01",
  end_observed = "2020-01-01",
  end_extrapolated = "2030-01-01"
)
```

# Ingesting Data

## LULC Data

For this tutorial, we do not read real observed LULC data:
instead, we generate a synthetic LULC raster with 3 layers (one per period) by adding autoregressive noise to a noisy raster.

```{r}
#| label: synthesize-and-agg-lulc-data
# autoregressive noise with skellam distribution
n_cells <- dim(template_rast)[1] * dim(template_rast)[2]
noise1 <- runif(n_cells, min = 0, max = 10)
noise2 <- noise1 + stats::rpois(n_cells, 1) - stats::rpois(n_cells, 1)
noise3 <- noise2 + stats::rpois(n_cells, 1) - stats::rpois(n_cells, 1)

synthetic_lulc <-
  rast(template_rast, nlyrs = 3, vals = c(noise1, noise2, noise3)) |>
  focal(w = 3, fun = mean, na.rm = TRUE) |>
  clamp(lower = 0, upper = 10) |>
  classify(rcl = data.frame(from = 0:9, to = 1:10, becomes = sample(1:10, 10)))
```

We now extract the generated values at our coordinates using `extract_using_coords_t`, giving us a tabular representation of `id_coord, id_period, src_class` tuples.
In a second step, we join in a long representation of the LULC metadata, associating `id_lulc` with `src_class`.

```{r}
synthetic_at_coords <- extract_using_coords_t(synthetic_lulc, db$coords_t)

synthetic_joint_meta <-
  synthetic_at_coords[, .(
    id_coord,
    id_period = substr(layer, 4, 4) |> as.integer(),
    src_class = value
  )][
    db$lulc_meta_long_v, # this is a long representation of id_lulc, src_class
    on = .(src_class),
    nomatch = NULL
  ]
```

We now have the LULC data in an almost canonical format.
We need to add information on which `id_run` this data belongs to: run 0 is the base run, providing observed data (see [`db$runs_t`](reference/runs_t) for details).
The call to [`as_lulc_data_t`](reference/lulc_data_t) ensures that the data we want to insert actually conforms to the form we are expecting.

```{r}
db$lulc_data_t <- as_lulc_data_t(synthetic_joint_meta[, .(
  id_run = 0L, # Base run ID
  id_period,
  id_lulc,
  id_coord
)])
```

Now that we have LULC data in our DB, we can derive data from it, e.g. we grab a view where a transition occurred, i.e. the anterior and posterior LULC ID are not the same.

```{r}
db$trans_v[id_lulc_anterior != id_lulc_posterior]
```

## Add Predictors and Neighbors

You have seen above how [`extract_using_coords_t`](reference/util_terra) can be used to transform a `SpatRaster` into a tabular form.
It can also be used with `SpatVector` objects, and the resulting tables need not just be LULC data: we could also use it to extract predictor information.
For demo purposes, we'll use test data that comes with `evoland`.
Note that the metadata holds a `fill_value` used to define what value should be used at a coordinate point with no explicit data set - e.g. if a square kilometre does not have a population count set, we can infer that it should be zero.

```{r}
#| label: add-predictors
db$pred_meta_t <- evoland:::test_pred_meta_t
db$pred_data_t <- evoland:::test_pred_data_t
```

Statistical models like GLMs or random forests lack inherent spatial concepts, so we explicitly calculate neighborhood relations for each coordinate to create spatial predictors (e.g., "number of neighboring forest cells within 100m").
This is done by creating a neighbor lookup table (`db$set_neighbors`) and then counting the number of neighbors within each distance break class and land use category (`db$generate_neighbor_predictors`).

```{r}
#| label: neighbors
db$set_neighbors(
  max_distance = 1000,
  distance_breaks = c(0, 100, 500, 1000)
)

db$generate_neighbor_predictors()
```

Have a look at the predictor metadata we have defined, it now contains new rows for the neighbor predictors:

```{r}
db$pred_meta_t
```

# Calibration

## Eligible Transitions and Predictor Pruning

We filter transitions eligible for modeling based on a minimum number of observed occurrences.

```{r}
#| label: eligible-transitions-and-predictors
db$trans_meta_t <- create_trans_meta_t(db$trans_v, min_cardinality_abs = 50)
```

Because each transition may be modelled using different predictors (aka *features*), we start out by setting the `trans_preds_t` table to the full cross product of viable transitions and predictors.
We then carry out a feature selection step in `get_pruned_trans_preds_t`, which here is used with a two-stage covariance filter.
First, candidate features are ranked and then selected up to a given correlation threshold, see [`covariance_filter`](reference/covariance_filter).
The assignment to `db$trans_preds_t` will overwrite the existing relations; you will be prompted if you want to do this if you're running R interactively.

```{r}
#| label: prune-trans-preds
#| warning: false
db$set_full_trans_preds()
# Warnings about non-convergence may occur due to our random synthetic data.
db$trans_preds_t <- db$get_pruned_trans_preds_t(
  filter_fun = covariance_filter,
  corcut = 0.1
)
```

## Transition Models

Now we fit partial models using training/validation splits, allowing for a goodness-of-fit (gof) estimation.
Here, we only fit a single partial model per transition, where we normally would create a list of candidates.
We can then pick the models with the best goodness of fit to retrain full models on all of the available predictor data; the full models are then used during extrapolation.

```{r}
#| label: trans-models
db$trans_models_t <- db$fit_partial_models(
  fit_fun = fit_glm,
  gof_fun = gof_glm,
  sample_frac = 0.7,
  seed = 42
)

db$trans_models_t <- db$fit_full_models(
  gof_criterion = "auc",
  gof_maximize = TRUE
)
```

## Transition Rates and Allocation Parameters

We estimate allocation parameters for DinamicaEGO, which determine the shape and size of new patches.

```{r}
#| label: transition-rates-alloc-params
db$trans_rates_t <-
  db$get_obs_trans_rates() |>
  extrapolate_trans_rates(
    periods = db$periods_t,
    coord_count = n_cells
  )

alloc_for_eval <- db$create_alloc_params_t(n_perturbations = 0)
alloc_for_eval[, id_run := 0L]
db$alloc_params_t <- alloc_for_eval
```

# Prediction + Allocation

With all components in place, we run the allocation step.
If Dinamica is not installed, you'll get a warning that the anterior LULC map is returned as the posterior.

```{r}
#| label: allocation
db$alloc_dinamica(
  id_period = db$periods_t[is_extrapolated == TRUE, id_period],
  gof_criterion = "auc",
  gof_maximize = TRUE
)
```

## Visualization

Finally, we can extract the simulated LULC maps and visualize them.

```{r}
#| label: visualization
labels <- db$periods_t[id_period != 0, paste0(year(start_date), " to ", year(end_date))]
plot_maps <- db$lulc_data_as_rast() |> setNames(labels)

plot(plot_maps)
```
