---
title: "Representing policy mechanisms in scenarios"
author: Jan Hartman
date: last-modified
vignette: >
  %\VignetteIndexEntry{policy-mechanisms}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
number-sections: true
---

```{r}
#| label: cleanup
#| include: false
unlink("policy-mechanisms.evolanddb", recursive = TRUE)

set.seed(666)
knitr::knit_hooks$set(seed = function(before, options, envir) {
  if (before && !is.null(options$seed)) {
    set.seed(options$seed)
  }
})
```

*This tutorial assumes the basic `evoland-plus` workflow from the introductory vignette and the
use of `runs_t` from the [stochastic allocation](stochastic-allocation-sensitivity.html)
vignette.*

## Policy mechanisms and where they act

Land use scenarios rarely differ in demand alone. Planning and policy also change *how* and
*where* land is converted, and some changes follow processes that no statistical transition
model captures. `evoland-plus` has no dedicated scenario machinery for this. Instead, each
mechanism is expressed as a scenario-specific version of one of the tables the simulation
already reads, written onto a run of its own:

| Mechanism | Examples | What it changes | Where it goes |
|---|---|---|---|
| Settlement form | compact development, densification, dispersed sprawl | the size and shape of newly converted patches | `alloc_params_t` for the scenario run |
| Spatial steering | building zones, nature conservation areas, agricultural priority areas | how likely each cell is to convert, before cells are allocated | `trans_pot_t` for the scenario run |
| Exogenous change | infrastructure projects, reservoirs, deglaciation | the allocated map itself, for processes outside the statistical model | `lulc_data_t` after allocation |

*How much* land changes is a fourth lever, the transition rates in `trans_rates_t`, and is
deliberately kept apart from the three above: they decide how and where a given demand is met,
not how large it is.

Readers of the SSP-CH scenarios will recognise these as the "pre-allocation", "allocation" and
"post-allocation" interventions there. The names used here describe the policy being
represented rather than the model step, because the step labels are misleading: potentials are
adjusted *before* anything is allocated.

We build one small synthetic study, allocate one period under a business-as-usual run, and then
apply each mechanism on a sibling run so its effect can be compared against that baseline.

## Synthetic study

The landscape, predictors, models and transition rates are exactly those of the
[stochastic allocation](stochastic-allocation-sensitivity.html) vignette, where each step is
explained; the code is folded here.

```{r}
#| fig-show: hide
#| code-fold: true
#| label: setup
#| output: false
library(evoland)
library(data.table)
library(terra)
```

```{r}
#| fig-show: hide
#| code-fold: true
#| label: create-db
#| collapse: true
db <- evoland_db$new(path = "policy-mechanisms.evolanddb")

db$lulc_meta_t <- create_lulc_meta_t(
  list(
    forest = list(pretty_name = "Forest"),
    arable = list(pretty_name = "Arable Land"),
    urban = list(pretty_name = "Urban Areas"),
    static = list(pretty_name = "Immutable")
  )
)

template_rast <- terra::rast(
  crs = "EPSG:2056",
  extent = terra::ext(c(
    xmin = 2697000,
    xmax = 2697000 + 30 * 100,
    ymin = 1252000,
    ymax = 1252000 + 30 * 100
  )),
  resolution = 100
)

db$coords_t <- create_coords_t_square(
  epsg = terra::crs(template_rast, describe = TRUE)$code |> as.integer(),
  extent = terra::ext(template_rast),
  resolution = terra::res(template_rast)[1]
)

db$periods_t <- create_periods_t(
  period_length_str = "P10Y",
  start_observed = "1995-01-01",
  end_observed = "2020-01-01",
  end_extrapolated = "2030-01-01"
)
```

```{r}
#| fig-show: hide
#| seed: 123
#| label: synthesize-lulc
#| code-fold: true
#| fig-asp: 0.3
n_cells <- dim(template_rast)[1] * dim(template_rast)[2]
noise1 <- runif(n_cells, min = -0.5, max = 4.2)
noise2 <- noise1 + stats::rpois(n_cells, 0.1) - stats::rpois(n_cells, 0.1)
noise3 <- noise2 + stats::rpois(n_cells, 0.1) - stats::rpois(n_cells, 0.1)

synthetic_lulc <-
  rast(template_rast, nlyrs = 3, vals = c(noise1, noise2, noise3)) |>
  focal(w = 3, fun = mean, na.rm = TRUE) |>
  clamp(lower = 0, upper = 4) |>
  classify(
    rcl = data.frame(from = 0:3, to = 1:4, becomes = c(2, 1, 3, 4)),
    right = FALSE
  ) |>
  setNames(paste0("synthetic LULC id_period=", 1:3))

plot(
  synthetic_lulc,
  nc = 3,
  col = data.frame(
    value = 1:4,
    color = c("#91B690", "#EB9486", "#F3DE8A", "#CBC6D2")
  )
)

db$lulc_data_t <-
  extract_using_coords_t(synthetic_lulc, db$coords_t)[,
    .(
      id_run = 0L,
      id_coord,
      id_period = substr(layer, 26, 26) |> as.integer(),
      id_lulc = value
    )
  ] |>
  as_lulc_data_t()
```

```{r}
#| fig-show: hide
#| label: predictors
#| code-fold: true
#| seed: 321
#| fig-asp: 0.3
# helper: scale raster to [0, 1]
scale01 <- function(x) {
  rng <- terra::global(x, c("min", "max"), na.rm = TRUE)
  (x - rng[1, 1]) / (rng[1, 2] - rng[1, 1])
}

# helper: smooth random field
smooth_field <- function(template, sd = 1, w = 7) {
  terra::setValues(template, rnorm(terra::ncell(template), sd = sd)) |>
    terra::focal(w = w, fun = mean, na.rm = TRUE) |>
    scale01()
}

# coordinate-based gradients
xy <- terra::crds(template_rast, df = TRUE)
x_grad <- terra::setValues(template_rast, (xy$x - min(xy$x)) / (max(xy$x) - min(xy$x)))
y_grad <- terra::setValues(template_rast, (xy$y - min(xy$y)) / (max(xy$y) - min(xy$y)))

latent1 <- smooth_field(template_rast, sd = 1, w = 9)
latent2 <- smooth_field(template_rast, sd = 1, w = 5)

accessibility <- scale01(0.55 * (1 - x_grad) + 0.25 * (1 - y_grad) + 0.20 * latent1)
site_quality <- scale01(0.50 * y_grad + 0.35 * latent2 + 0.15 * x_grad)

c(accessibility, site_quality) |>
  setNames(c("accessibility", "site_quality")) |>
  plot()
```

```{r}
#| fig-show: hide
#| label: predictors2
#| code-fold: true
#| seed: 321
#| fig-asp: 1
as_indicator <- function(x, class) {
  terra::app(x, fun = function(v) as.numeric(v %in% class))
}

forest_share <-
  as_indicator(synthetic_lulc, class = 1) |>
  terra::focal(w = 5, fun = mean, na.rm = TRUE) |>
  terra::mean()
arable_share <-
  as_indicator(synthetic_lulc, class = 2) |>
  terra::focal(w = 5, fun = mean, na.rm = TRUE) |>
  terra::mean()
urban_share <-
  as_indicator(synthetic_lulc, class = 3) |>
  terra::focal(w = 5, fun = mean, na.rm = TRUE)

make_dynamic <- function(base_context, static1, static2, noise_weight = 0.10) {
  out <- vector("list", terra::nlyr(base_context))
  for (i in seq_len(terra::nlyr(base_context))) {
    noise_i <- smooth_field(template_rast, sd = 1, w = 5)
    # fmt: skip
    out[[i]] <- scale01(
      0.55 * base_context[[i]] +
        0.25 * static1 +
        0.10 * static2 + noise_weight * noise_i
    )
  }
  terra::rast(out)
}

urban_pressure <-
  make_dynamic(
    base_context = urban_share,
    static1 = accessibility,
    static2 = 1 - site_quality
  )
add(urban_pressure) <-
  # fmt: skip
  scale01(
    0.60 * urban_pressure[[3]] +
      0.25 * urban_share[[3]] +
      0.10 * accessibility +
      0.05 * smooth_field(accessibility)
  )
names(urban_pressure) <-
  paste0("urban_pressure id_period=", 1:4)

forest_suitability <-
  make_dynamic(
    base_context = forest_share,
    static1 = site_quality,
    static2 = 1 - accessibility
  ) |>
  setNames("forest_suitability id_period=0")

arable_yield <-
  make_dynamic(
    base_context = arable_share,
    static1 = 1 - site_quality,
    static2 = accessibility
  ) |>
  setNames("arable_yield id_period=0")

# optional nuisance predictor that should usually score poorly
random_nuisance <-
  smooth_field(template_rast, sd = 1, w = 3) |>
  setNames("random_nuisance id_period=0")

plot(c(urban_pressure, forest_suitability, arable_yield, random_nuisance))
```

```{r}
#| fig-show: hide
#| label: predictors3
#| code-fold: true
#| seed: 321
#| fig-asp: 0.3
db$pred_meta_t <-
  create_pred_meta_t(list(
    urban_pressure = list(
      description = "synthetic confabulation",
      data_type = "float"
    ),
    forest_suitability = list(
      description = "synthetic confabulation",
      data_type = "float"
    ),
    arable_yield = list(
      description = "synthetic confabulation",
      data_type = "float"
    ),
    random_nuisance = list(
      description = "synthetic confabulation",
      data_type = "float"
    )
  ))

db$pred_data_t <-
  extract_using_coords_t(
    c(urban_pressure, forest_suitability, arable_yield, random_nuisance),
    db$coords_minimal
  )[, layer := as.character(layer)][,
    .(
      id_coord,
      id_run = 0L,
      id_period = substr(layer, nchar(layer), nchar(layer)) |> as.integer(),
      id_pred = fcase(
        grepl("urban_pressure", layer)     , 1L ,
        grepl("forest_suitability", layer) , 2L ,
        grepl("arable_yield", layer)       , 3L ,
        grepl("random_nuisance", layer)    , 4L
      ),
      value
    )
  ] |>
  as_pred_data_t()
```

```{r}
#| fig-show: hide
#| label: predictors4
#| code-fold: true
db$set_neighbors(
  max_distance = 1000,
  distance_breaks = c(0, 300, 1000),
  quiet = TRUE
)
db$generate_neighbor_predictors()
```

```{r}
#| fig-show: hide
#| code-fold: true
#| label: transitions-and-predictors
#| seed: 666
db$trans_meta_t <- create_trans_meta_t(
  db$trans_v,
  min_cardinality_abs = 20,
  exclude_anterior = 4 # exclude immutable class
)

db$set_full_trans_preds()

trans_pred_scored <- db$get_pred_filter_score(
  filter = mlr3filters::FilterImportance$new(
    learner = mlr3::lrn("classif.rpart")
  )
)

db$commit(
  trans_pred_scored[order(-importance)][, head(.SD, 3), by = id_trans], # top three per id_trans
  "trans_preds_t",
  method = "overwrite"
)
```

```{r}
#| fig-show: hide
#| code-fold: true
#| label: fit-models
#| seed: 666
trans_models <- db$fit_full_models(learner = mlr3::lrn("classif.rpart"))

# which transitions produced a usable model (non-null learner_full)? read this
# from the returned object before committing it
modeled_trans <- unique(
  trans_models$id_trans[!vapply(trans_models$learner_full, is.null, logical(1L))]
)

db$trans_models_t <- trans_models
```

```{r}
#| fig-show: hide
#| code-fold: true
#| label: reconcile-viable-transitions
trans_meta_reconciled <- db$trans_meta_t
trans_meta_reconciled[, is_viable := is_viable & id_trans %in% modeled_trans]
db$trans_meta_t <- trans_meta_reconciled
```

```{r}
#| fig-show: hide
#| code-fold: true
#| label: transition-rates
#| seed: 666
db$trans_rates_t <-
  # fmt: skip
  rowwiseDT(
    id_trans = , rate = ,
    1,           0.3,
    2,           0.2,
    3,           0.1,
    4,           0,
    5,           0,
    6,           0.2
  )[, `:=`(
    id_run = 0L,
    id_period = 4L,
    count = n_cells * rate
  )] |>
  as_trans_rates_t()

db$alloc_params_t <- db$create_alloc_params_t()
```

## Scenario runs

Every scenario is a child of the base run `0`, so it inherits the observed data, models,
transition rates and allocation parameters, and only overrides what its mechanism changes.

```{r}
#| label: register-runs
runs <- data.table(
  id_run = 0:4,
  parent_id_run = c(NA, 0L, 0L, 0L, 0L),
  description = c(
    "base",
    "business as usual",
    "compact settlement",
    "building zones and conservation",
    "reservoir project"
  )
)
db$commit(as_runs_t(runs), "runs_t", method = "overwrite")

lulc_ids <- db$lulc_meta_t[, setNames(id_lulc, name)]
urban_trans <- db$trans_meta_t[
  is_viable == TRUE & id_lulc_posterior == lulc_ids[["urban"]],
  id_trans
]
```

Transition potentials depend only on the anterior state and the predictors, which all runs
share, so we predict them once under the base run. Runs that do not edit them inherit these.

```{r}
#| label: predict-base
#| output: false
target_period <- 4L
db$id_run <- 0L
db$predict_trans_pot(
  id_period_post = target_period,
  select_score = "no.crossval",
  select_maximize = TRUE
)
```

Each run is allocated with `alloc_clumpy_one_period()`, which returns the map without
committing it. That leaves room for the map edits of the last mechanism; for the others,
`edit_map` does nothing. After committing, `upsert_new_neighbors()` refreshes the neighbourhood
predictors, which a following period would be predicted from. The same seed for every run keeps
the comparison about the mechanism rather than the draw.

```{r}
#| label: allocate-helper
allocate_run <- function(id_run, edit_map = identity) {
  db$id_run <- id_run
  set.seed(42)
  lulc <- alloc_clumpy_one_period(
    db = db,
    id_period_post = target_period,
    select_score = "no.crossval",
    select_maximize = TRUE
  )
  lulc <- edit_map(lulc)
  db$commit(lulc, "lulc_data_t", method = "upsert")
  db$upsert_new_neighbors(target_period)
  invisible(lulc)
}
```

## Business as usual

```{r}
#| label: allocate-bau
#| output: false
allocate_run(1L)
```

## Settlement form: compact development

A compact-development policy concentrates new building land in fewer, larger developments
instead of many scattered plots. In the allocation parameters this is the patch-size
distribution of transitions into urban land. For the scenario run we take the base estimate and
triple the mean size of new urban patches, scaling the variance so the spread of sizes grows
with it.

```{r}
#| label: settlement-form
compact <- copy(db$alloc_params_t)[, id_run := 2L]
compact[
  id_trans %in% urban_trans,
  `:=`(
    mean_patch_size = 3 * mean_patch_size,
    patch_size_variance = 9 * patch_size_variance
  )
]
db$alloc_params_t <- as_alloc_params_t(compact)
```

```{r}
#| label: allocate-compact
#| output: false
allocate_run(2L)
```

With Dinamica EGO as the allocator, the same table also carries the share of change that
expands existing patches (`frac_expander`); CLUMPY takes adjacency from the transition
potentials instead, through the neighbourhood predictors.

## Spatial steering: building zones and nature conservation

Zoning and protected areas change *where* land may convert. We represent them as masks on the
coordinates and adjust the transition potentials inside or outside them:

- **building zones**: new urban land is strongly preferred on land adjoining existing
  settlement (within 100 m, i.e. the neighbouring cells), so urban potentials outside that zone
  are scaled by 0.1;
- **nature conservation**: inside a protected area, no land converts to urban or arable use,
  so those potentials are set to zero.

```{r}
#| label: masks
lulc_anterior <- db$lulc_data_as_rast(id_period = target_period - 1L)

distance_to_urban <- terra::distance(
  terra::ifel(lulc_anterior == lulc_ids[["urban"]], 1, NA)
)
xy <- terra::crds(template_rast, df = TRUE)
protected_area <- terra::setValues(
  template_rast,
  (xy$x - 2698500)^2 + (xy$y - 1253500)^2 < 700^2
)

masks <- extract_using_coords_t(
  c(distance_to_urban <= 100, protected_area) |>
    setNames(c("building_zone", "protected_area")),
  db$coords_minimal
) |>
  dcast(id_coord ~ layer, value.var = "value")

plot(
  c(distance_to_urban <= 100, protected_area),
  main = c("Building zone", "Protected area"),
  legend = FALSE
)
```

The adjustments are multiplicative. That matters because allocation rescales each transition's
potentials to its demanded rate (see `adjusted_trans_pot_v()`): a factor states "k times as
likely here as elsewhere", which survives that rescaling, while an absolute value such as "set
to 0.1" would mean something different for every transition. A factor of zero is an exact
exclusion. It also means these mechanisms **relocate** change under a fixed demand rather than
adding or removing it; a policy that should change the amount belongs in the scenario's
transition rates.

The run inherits the base run's potentials slice by slice, one slice per transition and period.
We therefore write back the complete slice for every transition we touch, not only the masked
cells; a partial slice would hide the rest of the parent's.

```{r}
#| label: steer-potentials
db$id_run <- 3L
potentials <-
  db$trans_pot_t[id_period_post == target_period][
    masks,
    on = "id_coord",
    nomatch = NULL
  ][
    db$trans_meta_t[, .(id_trans, id_lulc_posterior)],
    on = "id_trans",
    nomatch = NULL
  ]

potentials[, factor := 1]
potentials[
  id_lulc_posterior == lulc_ids[["urban"]] & building_zone == 0,
  factor := 0.1
]
potentials[
  id_lulc_posterior %in% lulc_ids[c("urban", "arable")] & protected_area == 1,
  factor := 0
]

db$trans_pot_t <- as_trans_pot_t(potentials[, .(
  id_run = 3L,
  id_trans,
  id_period_post,
  id_coord,
  value = pmin(1, value * factor)
)])
```

`alloc_clumpy_one_period()` finds these potentials for run 3 and uses them instead of
predicting its own.

```{r}
#| label: allocate-steering
#| output: false
allocate_run(3L)
```

## Exogenous change: a reservoir project

Some change is decided outside the land market and has no counterpart in the observed
transitions the models learn from: a reservoir or road built to a plan, glacier retreat under a
climate scenario. These are applied to the allocated map directly. Here a planned reservoir
floods a strip of the landscape, turning it into immutable land.

```{r}
#| label: reservoir
reservoir <- db$coords_minimal[
  lat > 1252000 + 400 & lat < 1252000 + 900 & lon > 2697000 + 1200,
  id_coord
]

flood <- function(lulc) {
  lulc[id_coord %in% reservoir, id_lulc := lulc_ids[["static"]]]
  lulc
}
```

```{r}
#| label: allocate-reservoir
#| output: false
allocate_run(4L, edit_map = flood)
```

The flooded cells are extra change on top of the demanded rates. If the scenario's land
demand should already include them, subtract the project's conversions from that run's
`trans_rates_t` before allocating.

## Comparing the scenarios

```{r}
#| label: compare-maps
#| fig-width: 9
#| fig-asp: 0.85
db$id_run <- NULL
simulated <- db$fetch(
  "lulc_data_t",
  where = glue::glue("id_period = {target_period} and id_run between 1 and 4")
)
scenario_maps <- lapply(1:4, function(r) {
  tabular_to_raster(simulated[id_run == r], coords = db$coords_minimal)
}) |>
  terra::rast() |>
  setNames(c("Business as usual", "Compact settlement", "Zones + protection", "Reservoir"))

plot(
  scenario_maps,
  nc = 2,
  type = "classes",
  levels = db$lulc_meta_t[order(id_lulc), pretty_name],
  col = c("#91B690", "#EB9486", "#F3DE8A", "#CBC6D2"),
  mar = c(2, 2, 2, 7)
)
```

The table counts, per scenario, the cells that became urban, how many new urban patches they
form and their mean size, the share of new urban land inside the building zone, the conversions
to urban or arable land inside the protected area, and the cells flooded by the reservoir.

```{r}
#| label: compare-table
anterior <- db$fetch(
  "lulc_data_t",
  where = glue::glue("id_run = 0 and id_period = {target_period - 1L}")
)[, .(id_coord, id_lulc_anterior = id_lulc)]

changes <- simulated[anterior, on = "id_coord"][masks, on = "id_coord"][
  id_lulc != id_lulc_anterior
]
new_urban <- changes[id_lulc == lulc_ids[["urban"]]]

patch_stats <- rbindlist(lapply(1:4, function(r) {
  cells <- tabular_to_raster(
    new_urban[id_run == r, .(id_coord, value = 1)],
    coords = db$coords_minimal,
    value_col = "value"
  )
  sizes <- table(terra::values(terra::patches(cells, directions = 4)))
  data.table(id_run = r, new_patches = length(sizes), mean_patch_cells = mean(sizes))
}))

changes[,
  .(
    new_urban_cells = sum(id_lulc == lulc_ids[["urban"]]),
    share_in_building_zone = round(
      mean(building_zone[id_lulc == lulc_ids[["urban"]]] == 1),
      2
    ),
    converted_in_protected_area = sum(
      protected_area == 1 & id_lulc %in% lulc_ids[c("urban", "arable")]
    ),
    flooded_cells = sum(id_lulc == lulc_ids[["static"]])
  ),
  keyby = id_run
][patch_stats, on = "id_run"][
  runs[, .(id_run, scenario = description)],
  on = "id_run",
  nomatch = NULL
]
```

Compared with business as usual:

- **compact settlement** converts the same amount of land in far fewer, larger patches;
- **building zones and conservation** keep a similar quantity of new urban land but move more of
  it into the zone, and nothing converts to urban or arable land inside the protected area any
  more; the quantity is not exactly equal because CLUMPY's single-cell allocation meets the
  demanded rates in expectation;
- **the reservoir project** floods its strip regardless of what the allocator did there,
  overwriting some cells it had just converted to urban. In a following period, the
  neighbourhood predictors would see the reservoir.

## Summary

| Mechanism | Scenario-specific table | When it is applied |
|---|---|---|
| Settlement form | `alloc_params_t` | written once for the run |
| Spatial steering | `trans_pot_t` | after prediction, before allocation, per period |
| Exogenous change | `lulc_data_t` | after allocation, before committing, per period |

All three rely on the run lineage in `runs_t` to override only what differs from the parent
run, and the per-period ones on the single-period allocators
(`alloc_clumpy_one_period()`, `alloc_dinamica_one_period()`), which leave the loop to the
caller.
