---
title: "Stochastic allocation sensitivity with `runs_t`"
author: Jan Hartman
date: last-modified
vignette: >
  %\VignetteIndexEntry{stochastic-allocation-sensitivity}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
number-sections: true
---

```{r}
#| label: cleanup
#| include: false
unlink("stochastic-sensitivity.evolanddb", recursive = TRUE)

set.seed(666)
knitr::knit_hooks$set(seed = function(before, options, envir) {
  if (before && !is.null(options$seed)) {
    set.seed(options$seed)
  }
})

local({
  chunk_messages <- character()

  knitr::knit_hooks$set(message = function(x, options) {
    chunk_messages <<- c(chunk_messages, x)
    return(character(0))
  })

  default_chunk_hook <- knitr::knit_hooks$get("chunk")
  knitr::knit_hooks$set(chunk = function(x, options) {
    out <- default_chunk_hook(x, options)
    if (length(chunk_messages) > 0) {
      msg_text <- paste(chunk_messages, collapse = "")
      callout <- paste0(
        "\n\n<details><summary>Messages</summary><pre>\n",
        msg_text,
        "</pre></details>\n"
      )
      chunk_messages <<- character()
      out <- paste0(out, callout)
    }
    return(out)
  })
})
```

*Note: this tutorial assumes that you already know the basic `evoland-plus` workflow and the concepts of transition potentials, constrained demand, and patch-based allocation. The preceding tutorial introduced those ideas; here we focus on some advantages of evoland-style sensitivity analyses.*

## Why another tutorial?

The default tutorial walks through a single calibration-and-allocation workflow, ending with one extrapolated realization. That is a good first contact, but it hides an important property of patch-based allocators: **once transition demand and transition potentials are fixed, the final pattern can still vary from run to run because allocation is stochastic**.

For sensitivity analysis, a single map is therefore not enough. In this vignette we will:

1. establish a simple synthetic model,
2. register 30 allocation realizations in `runs_t`,
3. allocate exactly one extrapolated period 30 times,
4. compare each extrapolated realization to the last observed map, and
5. summarize the results as a **change-frequency heatmap** showing where change happens consistently.

This is useful whenever you want to distinguish **structurally robust change**, i.e. cells that nearly always switch class.


## Setup

```{r}
#| label: setup
#| output: false
library(evoland)
library(data.table)
library(terra)
```

We create a fresh on-disk database with some synthetic land use categories;
we use a spatial domain of 30x30 cells on a square grid defined in meters and retain a template raster for later use;
and we set up three "observed" periods to be filled with synthetic data, plus a single extrapolated period .

```{r}
#| label: create-db
#| collapse: true
db <- evoland_db$new(path = "stochastic-sensitivity.evolanddb")

db$lulc_meta_t <- create_lulc_meta_t(
  list(
    forest = list(pretty_name = "Forest"),
    arable = list(pretty_name = "Arable Land"),
    urban = list(pretty_name = "Urban Areas"),
    static = list(pretty_name = "Immutable")
  )
)

template_rast <- terra::rast(
  crs = "EPSG:2056",
  extent = terra::ext(c(
    xmin = 2697000,
    xmax = 2697000 + 30 * 100,
    ymin = 1252000,
    ymax = 1252000 + 30 * 100
  )),
  resolution = 100
)

db$coords_t <- create_coords_t_square(
  epsg = terra::crs(template_rast, describe = TRUE)$code |> as.integer(),
  extent = terra::ext(template_rast),
  resolution = terra::res(template_rast)[1]
)

db$periods_t <- create_periods_t(
  period_length_str = "P10Y",
  start_observed = "1995-01-01",
  end_observed = "2020-01-01",
  end_extrapolated = "2030-01-01"
)
```

## Synthetic observed data

As in the introductory vignette, we create a small autoregressive synthetic landscape with three observed raster layers.

```{r}
#| seed: 123
#| label: synthesize-lulc
#| fig-asp: 0.3
n_cells <- dim(template_rast)[1] * dim(template_rast)[2]
noise1 <- runif(n_cells, min = -0.5, max = 4.2)
noise2 <- noise1 + stats::rpois(n_cells, 0.1) - stats::rpois(n_cells, 0.1)
noise3 <- noise2 + stats::rpois(n_cells, 0.1) - stats::rpois(n_cells, 0.1)

synthetic_lulc <-
  rast(template_rast, nlyrs = 3, vals = c(noise1, noise2, noise3)) |>
  focal(w = 3, fun = mean, na.rm = TRUE) |>
  clamp(lower = 0, upper = 4) |>
  classify(
    rcl = data.frame(from = 0:3, to = 1:4, becomes = c(2, 1, 3, 4)),
    right = FALSE
  ) |>
  setNames(paste0("synthetic LULC id_period=", 1:3))

plot(
  synthetic_lulc,
  nc = 3,
  col = data.frame(
    value = 1:4,
    color = c("#91B690", "#EB9486", "#F3DE8A", "#CBC6D2")
  )
)
```

We now transform the raster stack into `lulc_data_t` records and attach `id_run = 0L`, which is the reserved base run for observed data.

```{r}
#| label: ingest-lulc
db$lulc_data_t <-
  extract_using_coords_t(synthetic_lulc, db$coords_t)[,
    .(
      id_coord,
      id_period = substr(layer, 26, 26) |> as.integer(),
      id_lulc = value
    )
  ][,
    .(id_run = 0L, id_period, id_lulc, id_coord)
  ] |>
  as_lulc_data_t()
```

## Predictors and neighborhoods

Here, we confabulate some predictors based on gradient based "latent" influences then used to construct an "accessibility" and "site quality" quantifier.
These are then used to influence some variables that are somewhat correlated with the synthetic land use map we constructed earlier.

```{r}
#| label: predictors
#| collapse: true
#| seed: 321
# helper: scale raster to [0, 1]
scale01 <- function(x) {
  rng <- terra::global(x, c("min", "max"), na.rm = TRUE)
  (x - rng[1, 1]) / (rng[1, 2] - rng[1, 1])
}

# helper: smooth random field
smooth_field <- function(template, sd = 1, w = 7) {
  terra::setValues(template, rnorm(terra::ncell(template), sd = sd)) |>
    terra::focal(w = w, fun = mean, na.rm = TRUE) |>
    scale01()
}

# coordinate-based gradients
xy <- terra::crds(template_rast, df = TRUE)
x_grad <- terra::setValues(template_rast, (xy$x - min(xy$x)) / (max(xy$x) - min(xy$x)))
y_grad <- terra::setValues(template_rast, (xy$y - min(xy$y)) / (max(xy$y) - min(xy$y)))

latent1 <- smooth_field(template_rast, sd = 1, w = 9)
latent2 <- smooth_field(template_rast, sd = 1, w = 5)

accessibility <- scale01(0.55 * (1 - x_grad) + 0.25 * (1 - y_grad) + 0.20 * latent1)
site_quality <- scale01(0.50 * y_grad + 0.35 * latent2 + 0.15 * x_grad)

plot(c(accessibility, site_quality))

as_indicator <- function(x, class) {
  terra::app(x, fun = function(v) as.numeric(v %in% class))
}

forest_share <-
  as_indicator(synthetic_lulc, class = 1) |>
  terra::focal(w = 5, fun = mean, na.rm = TRUE) |>
  terra::mean()
arable_share <-
  as_indicator(synthetic_lulc, class = 2) |>
  terra::focal(w = 5, fun = mean, na.rm = TRUE) |>
  terra::mean()
urban_share <-
  as_indicator(synthetic_lulc, class = 3) |>
  terra::focal(w = 5, fun = mean, na.rm = TRUE)

make_dynamic <- function(base_context, static1, static2, noise_weight = 0.10) {
  out <- vector("list", terra::nlyr(base_context))
  for (i in seq_len(terra::nlyr(base_context))) {
    noise_i <- smooth_field(template_rast, sd = 1, w = 5)
    # fmt: skip
    out[[i]] <- scale01(
      0.55 * base_context[[i]] +
        0.25 * static1 +
        0.10 * static2 + noise_weight * noise_i
    )
  }
  terra::rast(out)
}

urban_pressure <-
  make_dynamic(
    base_context = urban_share,
    static1 = accessibility,
    static2 = 1 - site_quality
  )
add(urban_pressure) <-
  # fmt: skip
  scale01(
    0.60 * urban_pressure[[3]] +
      0.25 * urban_share[[3]] +
      0.10 * accessibility +
      0.05 * smooth_field(accessibility)
  )
names(urban_pressure) <-
  paste0("urban_pressure id_period=", 1:4)

forest_suitability <-
  make_dynamic(
    base_context = forest_share,
    static1 = site_quality,
    static2 = 1 - accessibility
  ) |>
  setNames("forest_suitability id_period=0")

arable_yield <-
  make_dynamic(
    base_context = arable_share,
    static1 = 1 - site_quality,
    static2 = accessibility
  ) |>
  setNames("arable_yield id_period=0")


# optional nuisance predictor that should usually score poorly
random_nuisance <-
  smooth_field(template_rast, sd = 1, w = 3) |>
  setNames("random_nuisance id_period=0")

plot(c(urban_pressure, forest_suitability, arable_yield, random_nuisance))

db$pred_meta_t <-
  create_pred_meta_t(list(
    urban_pressure = list(
      description = "synthetic confabulation",
      data_type = "float"
    ),
    forest_suitability = list(
      description = "synthetic confabulation",
      data_type = "float"
    ),
    arable_yield = list(
      description = "synthetic confabulation",
      data_type = "float"
    ),
    random_nuisance = list(
      description = "synthetic confabulation",
      data_type = "float"
    )
  ))

db$pred_data_t <-
  extract_using_coords_t(
    c(urban_pressure, forest_suitability, arable_yield, random_nuisance),
    db$coords_minimal
  )[, layer := as.character(layer)][,
    .(
      id_coord,
      id_run = 0L,
      id_period = substr(layer, nchar(layer), nchar(layer)) |> as.integer(),
      id_pred = fcase(
        grepl("urban_pressure", layer)     , 1L ,
        grepl("forest_suitability", layer) , 2L ,
        grepl("arable_yield", layer)       , 3L ,
        grepl("random_nuisance", layer)    , 4L
      ),
      value
    )
  ] |>
  as_pred_data_t()

db$set_neighbors(
  max_distance = 1000,
  distance_breaks = c(0, 300, 1000),
  quiet = TRUE
)
db$generate_neighbor_predictors()
```

## Calibration

### Eligible transitions and predictor pruning

```{r}
#| label: transitions-and-predictors
#| seed: 666
db$trans_meta_t <- create_trans_meta_t(
  db$trans_v,
  min_cardinality_abs = 20,
  exclude_anterior = 4 # exclude immutable class
)
# db$trans_meta_t <-
# data.table::rowwiseDT(
#   id_trans=, id_lulc_anterior=, id_lulc_posterior=, cardinality=, frequency_rel=, frequency_abs=, is_viable=,
#  4,                1,                 2,          13,   0.061032864,  0.0072222222,     FALSE,
#  3,                1,                 3,          81,   0.380281690,  0.0450000000,      TRUE,
#  1,                2,                 1,          23,   0.107981221,  0.0127777778,      TRUE,
#  2,                3,                 1,          75,   0.352112676,  0.0416666667,      TRUE,
#  5,                3,                 4,          20,   0.093896714,  0.0111111111,      TRUE,
#  6,                4,                 3,           1,   0.004694836,  0.0005555556,     FALSE
# ) |> as_trans_meta_t()

db$set_full_trans_preds()

trans_pred_scored <- db$get_pred_filter_score(
  filter = mlr3filters::FilterImportance$new(
    learner = mlr3::lrn("classif.rpart")
  )
)

db$commit(
  trans_pred_scored[order(-importance)][, head(.SD, 3), by = id_trans], # top three per id_trans
  "trans_preds_t",
  method = "overwrite"
)
```

### Transition models

```{r}
#| label: fit-models
#| collapse: true
#| seed: 666
db$trans_models_t <- db$fit_partial_models(
  learner = mlr3::lrn("classif.rpart"),
  measures = c("classif.auc", "classif.acc"),
  sample_frac = 0.7,
  seed = 666
)

db$trans_models_t <- db$fit_full_models(
  select_score = "classif.auc",
  select_maximize = TRUE
)
```

### Transition rates and baseline allocation parameters

```{r}
#| label: transition-rates
#| seed: 666
db$trans_rates_t <-
  # fmt: skip
  rowwiseDT(
    id_trans = , rate = ,
    1,           0.3,
    2,           0.2,
    3,           0.1,
    4,           0,
    5,           0,
    6,           0.2
  )[, `:=`(
    id_run = 0L,
    id_period = 4L,
    count = n_cells * rate
  )] |>
  as_trans_rates_t()

db$alloc_params_t <-
  db$create_alloc_params_t(n_perturbations = 0)[, id_run := 0L]
```

## Registering stochastic realizations in `runs_t`

The central idea of this tutorial is that **each stochastic allocation is a separate run**. We therefore create 30 run records and propagate the same baseline allocation parameters to each run.

Because the exact metadata fields of `runs_t` may evolve over time, the helper below only assumes that `id_run` exists and fills other common metadata columns when present.

```{r}
#| label: register-runs
n_realizations <- 30L
seed_start <- 1000L
runs_new <- data.table(
  id_run = seq_len(n_realizations + 1) - 1L,
  parent_id_run = NA_integer_,
  description = "base",
  kind = "synthetic 'historical' case",
  seed = 666L
)
runs_new[id_run > 0, id_run := seq_len(n_realizations)]
runs_new[id_run > 0, parent_id_run := 0L]
runs_new[id_run > 0, description := paste("stochastic allocation", id_run)]
runs_new[id_run > 0, kind := "allocation_sensitivity"]
runs_new[id_run > 0, seed := id_run + seed_start]

db$commit(as_runs_t(runs_new), "runs_t", method = "overwrite")

runs_out <- db$runs_t[id_run > 0]
run_ids <- runs_out$id_run
```

We now duplicate the baseline allocation parameters for every stochastic run. This ensures that **all differences among runs arise from the allocator's stochasticity, not from parameter perturbations**.

## Running one extrapolated period 30 times

We only allocate the first extrapolated period. This keeps the comparison clean: every realization starts from the same observed state.

```{r}
#| label: identify-periods
last_observed_id <- db$periods_t[
  id_period != 0 & is_extrapolated == FALSE,
  max(id_period)
]

first_extrapolated_id <- db$periods_t[
  is_extrapolated == TRUE,
  min(id_period)
]

last_observed_id
first_extrapolated_id
```

The wrapper below is intentionally defensive. Some versions of the API may expose `id_run` directly in `alloc_clumpy()`, while others may infer runs from the database state. If the method has an `id_run` argument, we pass it; otherwise we fall back to the method's default signature.

```{r}
#| label: allocation-wrapper
.mapply(
  dots = runs_out,
  FUN = function(id_run, seed, ...) {
    db$id_run <- id_run
    db$alloc_clumpy(
      id_period = 4, # single extrapolated period
      select_score = "classif.auc",
      select_maximize = TRUE,
      seed = seed
    )
  },
  MoreArgs = NULL
)
```

## Summarizing stochastic sensitivity

A convenient first summary is binary: for each cell we ask whether its simulated class differs from the last observed class. Averaging that binary indicator over 30 realizations gives a **change frequency** between 0 and 1.

- `0` means the cell never changes.
- `1` means the cell always changes.
- intermediate values indicate uncertainty induced by the allocator.

```{r}
#| label: summarize-change-frequency
db$id_run <- NULL

# 900 locations x 1 observation = 900 rows
observed_last <- db$fetch("lulc_data_t", where = "id_run = 0 and id_period = 3")[,
  .(id_coord, id_lulc_observed = id_lulc)
]

# 900 locations * 30 runs = 27000 rows
simulated_extrap <- db$fetch("lulc_data_t", where = "id_run > 0 and id_period = 4")[,
  .(id_run, id_coord, id_lulc_simulated = id_lulc)
]

change_maps <-
  simulated_extrap[
    observed_last,
    on = .(id_coord)
  ][,
    .(
      n_changed = sum(id_lulc_simulated != id_lulc_observed) |> as.double(),
      p_changed = mean(id_lulc_simulated != id_lulc_observed),
      observed = id_lulc_observed[1] |> as.double(),
      mode_simulated = id_lulc_simulated |> tabulate() |> which.max() |> as.double()
    ),
    by = .(id_coord)
  ] |>
  data.table::melt(id.vars = "id_coord") |>
  tabular_to_raster(coords = db$coords_minimal, value_col = "value")
```

To visualize the result, we join coordinates back in and rasterize the per-cell change frequency.


## Heatmap of stochastic change consistency

```{r}
#| label: plot-change-frequency
#| fig-width: 7
#| fig-height: 6
plot(
  change_maps$variable_p_variable_changed,
  main = "Share of runs in which a cell changes class"
)
```

The interpretation is straightforward:

- **cooler / lighter cells** are stable under repeated allocation,
- **warmer / darker cells** change consistently and may reflect strong transition potential and/or limited competition, and
- **mid-range cells** reveal locations where several cells compete for a limited transition demand and small stochastic differences decide which ones actually convert.

In other words, this map is not a transition-potential surface. It is a **realization-frequency surface**.

## Inspecting a few individual realizations

The aggregate heatmap is informative, but it helps to also inspect a few single runs side by side.

We can have even more fun with these visualisations: here's an animation of each realisation

```{r}
simulated_extrap |>
  tabular_to_raster(coords = db$coords_minimal, value_col = "id_lulc_simulated") |>
  animate(
    pause = 0.5,
    legend = FALSE,
    axes = FALSE,
    box = FALSE,
    plg = list(),
    mar = c(0.1, 0.1, 0.1, 0.1),
    main = "Realisations of stochastic allocation",
    col = data.frame(
      value = 1:4,
      color = c("#91B690", "#EB9486", "#F3DE8A", "#CBC6D2")
    )
  )
```

Even with identical calibration, transition rates, and allocation parameters, local differences appear across runs. The heatmap above compresses those differences into a robust summary.

## Optional: class-specific consistency maps

The previous heatmap asked only whether a cell changes at all. You can build more targeted summaries in the same way. For example, the next chunk computes the frequency with which a cell becomes `urban`.

```{r}
#| label: urban-frequency
urban_frequency <-
  simulated_extrap[,
    .(
      p_urban = mean(id_lulc_simulated == 3L) #urban id
    ),
    by = .(id_coord)
  ] |>
  tabular_to_raster(db$coords_minimal, value_col = "p_urban")

plot(
  urban_frequency,
  main = "Share of runs in which a cell ends up urban",
)
```

This is often the more policy-relevant view if stakeholders care about a specific target class rather than generic change.

## What `runs_t` buys you

In a small tutorial like this, we could have stored realizations in ad hoc objects. Using `runs_t` is better because it scales to more serious experiments:

- each realization is explicitly registered,
- run-level metadata such as seeds, scenario labels, or parameter sets can be stored alongside the simulation,
- downstream summaries can be grouped by run families, and
- the same database structure can later support sensitivity analysis, parameter sweeps, or scenario ensembles.

A natural next step would be to combine **between-run stochasticity** (shown here) with **between-parameter variability** by creating multiple allocation parameter sets per scenario and multiple stochastic replicates per parameter set.

## Take-away

Repeated allocation changes the interpretation of the model output:

- one extrapolated map is a **draw**,
- the stack of 30 realizations is an **ensemble**, and
- the heatmap of change frequency is a compact way to communicate **where the ensemble is robust and where it is uncertain**.

That makes `runs_t` more than bookkeeping: it becomes the backbone for uncertainty-aware land use change analysis.
